---
title: "Lending Club data analysis"
author: "Krzysztof Lis"
date: "5 Aug 2016"
---

```{r load, echo=FALSE, warning=FALSE, message=FALSE}
library(choroplethr)
library(choroplethrMaps)
library(DescTools)
library(dplyr)
library(ggplot2)
library(readr)
library(acs)

options(scipen=999)
LOANS_DATA <- read_csv('lc_data_processed.csv')
spec(LOANS_DATA)
LOANS_DATA <- subset(LOANS_DATA,annual_inc < 250000)
#correct dti and annual_inc data
LOANS_DATA$dti = ifelse(LOANS_DATA$dti > 100, 100, LOANS_DATA$dti)
LOANS_DATA$annual_inc[is.na(LOANS_DATA$annual_inc)] <- median(LOANS_DATA$annual_inc, na.rm = TRUE)
LOANS_DATA$paid_amount_ratio <- LOANS_DATA$total_rec_prncp/LOANS_DATA$loan_amnt
LOANS_DATA$own_house <- revalue(LOANS_DATA$home_ownership, c("MORTGAGE" = 0, "RENT" = 0, "OTHER" = 0, "NONE" = 0, "ANY" = 0, "OWN" = 1))

# load the state names
data(state.regions)

# merge the loan book with the state names
LOANS_DATA <- merge(LOANS_DATA, state.regions, by.x = "addr_state", by.y = "abb")

```

## Maps

### Amount by state
```{r, echo=TRUE}
amount_by_state <-
LOANS_DATA %>% group_by(region) %>%
  summarise(value = sum(loan_amnt, na.rm=TRUE))

state_choropleth(amount_by_state, title = "Value by State")
```

### Volume by state
```{r, echo=TRUE}
volume_by_state <-
LOANS_DATA %>% group_by(region) %>%
  summarise(value = n())

state_choropleth(volume_by_state, title = "Volume by State")
```

## Loan amounts

I focused on several variables and I plotted them using DescTools package.
Here I created the density plot, box plot, and empirical distribution function plot.
```{r, echo=TRUE}
Desc(LOANS_DATA$loan_amnt, main = "Loan amount distribution", plotit = TRUE)
```
The annual income:
```{r, echo=TRUE}
Desc(LOANS_DATA$annual_inc, main = "Annual income distribution", plotit = TRUE)

```
Here's how the loan book was growing:
```{r, echo=TRUE}
LOANS_DATA$issue_d <- as.Date(gsub("^", "01-", LOANS_DATA$issue_d), format="%d-%b-%Y")

amnt_df <- LOANS_DATA %>% 
  select(issue_d, loan_amnt) %>% 
  group_by(issue_d) %>% 
  summarise(Amount = sum(loan_amnt))

ts_amnt <- ggplot(amnt_df, 
                  aes(x = issue_d, y = Amount))
ts_amnt + geom_line() + xlab("Date issued")
```

## Loan reasons

```{r, echo=TRUE}
Desc(LOANS_DATA$purpose, main = "Loan purposes", plotit = TRUE)

```
## Loan grades


```{r, echo=TRUE}
Desc(LOANS_DATA$grade, main = "Loan grades", plotit = TRUE)
```

### Interest rate ~ grade
```{r, echo=TRUE}
Desc(int_rate ~ grade, LOANS_DATA, digits = 1, main = "Interest rate by grade", plotit = TRUE)
```

## Loan statuses

```{r, echo=TRUE}
Desc(LOANS_DATA$loan_status, plotit = T)
```

Then I checked the distribution of loan amounts by status.
```{r, echo=TRUE}
box_status <- ggplot(LOANS_DATA, aes(loan_status, loan_amnt))
box_status + geom_boxplot(aes(fill = loan_status)) +
  theme(axis.text.x = element_blank()) +
  labs(list(
    title = "Loan amount by status",
    x = "Status",
    y = "Amount"))  
```

Here's how the value of loans of different grades was changing over time
```{r, echo=TRUE}
amnt_df_grade <- LOANS_DATA %>% 
  select(issue_d, loan_amnt, grade) %>% 
  group_by(issue_d, grade) %>% 
  summarise(Amount = sum(loan_amnt))

ts_amnt_grade <- ggplot(amnt_df_grade, 
                  aes(x = issue_d, y = Amount))
ts_amnt_grade + geom_area(aes(fill=grade)) + xlab("Date issued")
```

## Numeric variables

A correlation matrix was calculated for numeric variables in order to remove redundant data (with correlation coefficient grater than 90%):
```{r, echo=TRUE, warning=FALSE, message=FALSE}
set.seed(123)
library(mlbench)
library(caret)

numeric_vars <- sapply(LOANS_DATA, is.numeric)
correlationMatrix <- cor(LOANS_DATA[, numeric_vars])
correlationMatrix[is.na(correlationMatrix)] <- 0
redundantData <- findCorrelation(correlationMatrix, cutoff = 0.9, verbose = TRUE, names = TRUE, exact = TRUE)
print(correlationMatrix[redundantData, redundantData])
```

The impact of numerical variables on loans beahaviour:

###Loan amount
```{r, echo=TRUE, warning=FALSE, message=FALSE}
ggplot(data=LOANS_DATA, aes(x=LOANS_DATA$loan_amnt, colour=factor(defaulted))) + geom_density()
```

###Interest rates
```{r, echo=TRUE, warning=FALSE, message=FALSE}
ggplot(data=LOANS_DATA, aes(x=LOANS_DATA$int_rate, colour=factor(defaulted))) + geom_density()
```

###Term
```{r, echo=TRUE, warning=FALSE, message=FALSE}
ggplot(data=LOANS_DATA, aes(x=LOANS_DATA$term, colour=factor(defaulted))) + geom_density()
```

###Employment length
```{r, echo=TRUE, warning=FALSE, message=FALSE}
ggplot(data=LOANS_DATA, aes(x=LOANS_DATA$emp_length, colour=factor(defaulted))) + geom_density()
```

###Annual income
```{r, echo=TRUE, warning=FALSE, message=FALSE}
ggplot(data=LOANS_DATA, aes(x=LOANS_DATA$annual_inc, colour=factor(defaulted))) + geom_density()
```

###Delinque
```{r, echo=TRUE, warning=FALSE, message=FALSE}
ggplot(data=LOANS_DATA, aes(x=LOANS_DATA$delinq_2yrs, colour=factor(defaulted))) + geom_density() + xlim(0,10)
```

###Debt to income
```{r, echo=TRUE, warning=FALSE, message=FALSE}
ggplot(data=LOANS_DATA, aes(x=LOANS_DATA$dti, colour=factor(defaulted))) + geom_density() + xlim(0,50)
```

###Paid amount ratio
```{r, echo=TRUE, warning=FALSE, message=FALSE}
ggplot(data=LOANS_DATA, aes(x=LOANS_DATA$paid_amount_ratio, colour=factor(defaulted))) + geom_density()
```